---
title: "functions and packages affected by deprecating rgdal, rgeos and maptools"
date: "7 December 2022"
format: html
---

The reason for these listings this is to find and anticipate what will happen when [rgdal, rgeos and maptools will retire](https://r-spatial.org/r/2022/04/12/evolution.html). Possibilities to migrate to more modern functions in package `sf` are given [here](https://github.com/r-spatial/sf/wiki/Migrating).

## `pkgapi` analysis

The tables below, created from analysing and filtering the output from running `pkgapi::map_package()` on all packages meeting `tools::package_dependencies(packages = <pkg>, tools::available.packages(), which = "most", recursive = FALSE, reverse = TRUE)`, where `<pkg>` are `maptools`, `rgeos` and `rgdal` each run separately, list:

* which functions in packages `maptools`, `rgeos` and `rgdal` are used by which packages, ordered by the number of packages affected, and 
* the functions in `maptools`, `rgeos` and `rgeos` that are called by each package involved, ordered by the number of functions involved

Note that this only concerns _exported functions_, not necessarily S3 or S4 methods, and so gives an under estimation of the amount of dependency on the three packages.


### By deprecating function

```{r}
r = read.csv("pkgapi_221206.csv", header = FALSE)
names(r) = c("function", "packages")
r$packages = lengths(strsplit(r[[2]], " "))
r_o = r[order(r$packages, decreasing = TRUE),]
library(knitr)
kable(r_o, row.names = FALSE)
```

### By depending package

```{r}
r = read.csv("pkgapi_by_pkg_221206.csv", header = FALSE)
names(r) = c("package", "functions involved")
r$functions = lengths(strsplit(r$`functions involved`, " ")) # - 1 # leading space
kable(r[order(r$functions, decreasing = TRUE),c(3,1,2)], row.names = FALSE)
```

### Impending changes in `raster`

Many packages using `raster` declare `rgeos` and/or `rgdal` as required in some form, but very recent changes in `raster`, removing both these packages entirely from `DESCRIPTION`, will in time propagate, so that declarations of these packages in packages really only needing `raster` can be deleted soon.


